---
output:
  github_document:
    html_preview: false
---

<!-- DEVELOPERS.md is generated from DEVELOPERS.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Notes for developers

## Requirements

- R (>= 4.0), [just](https://github.com/casey/just), [air](https://github.com/posit-dev/air/), [Quarto CLI](https://quarto.org/docs/get-started/)

Run `just list` to see all available recipes.

## Data pipeline

Users call `get_*()`, which reads `tidy_data_sources.yml` for the current version and URL, downloads the pre-built parquet if not already cached, and caches it to `~/.cache/R/nhstt/tidy/{dataset}.parquet`. Subsequent calls return the cached file instantly.

Developers run `build_tidy_data(dataset)` (`just build-data` for all), which downloads raw source files, applies the tidy pipeline for every period, and writes a combined parquet to `data-raw/`. Parquets are published as GitHub Release assets and are directly usable from Python, Julia, or any language that reads parquet.

### Available datasets

```{r datasets, echo = FALSE, message = FALSE}
df <- nhstt:::available_nhstt_reports() |>
  dplyr::select(
    get_function,
    frequency,
    first_period,
    last_period,
    n_periods,
    version
  ) |>
  dplyr::mutate(get_function = paste0("`", get_function, "()`")) |>
  dplyr::distinct()

knitr::kable(
  df,
  row.names = FALSE,
  col.names = c(
    "Function",
    "Frequency",
    "First period",
    "Last period",
    "Periods",
    "Version"
  ),
  align = c("l", "l", "l", "l", "r", "r")
)
```

### Approximate build times

One period timed per dataset; extrapolated to full build. Re-run with `just render-developers`.

```{r build-times, echo = FALSE, message = FALSE}
reports <- nhstt:::available_nhstt_reports() |>
  dplyr::filter(!stringr::str_detect(title, "metadata")) |>
  dplyr::distinct(dataset, frequency, last_period, n_periods)

timings <- lapply(seq_len(nrow(reports)), function(i) {
  d <- reports[i, ]
  elapsed <- system.time(
    nhstt:::download_and_tidy(d$dataset, d$last_period, d$frequency)
  )[["elapsed"]]
  data.frame(
    dataset = paste0("`", d$dataset, "`"),
    timed_period = d$last_period,
    n_periods = d$n_periods,
    time_one_s = round(elapsed, 1),
    est_total_min = round(elapsed * d$n_periods / 60, 1)
  )
})

knitr::kable(
  do.call(rbind, timings),
  row.names = FALSE,
  col.names = c(
    "Dataset",
    "Timed period",
    "Periods",
    "Time for one (s)",
    "Est. full build (min)"
  ),
  align = c("l", "l", "r", "r", "r")
)
```

## Publishing new data

Each dataset is released independently — updating monthly data does not affect annual dataset versions or caches.

1. Update raw config (e.g. `raw_monthly_data_config.yml`) with new sources
2. `just build-data` — rebuilds all parquets and writes to `data-raw/`
3. Create a GitHub Release with a dataset-specific tag and upload that parquet:
   ```bash
   gh release create activity-performance-monthly-v0.4.0 \
     data-raw/activity_performance_monthly.parquet \
     --title "activity-performance-monthly v0.4.0"
   ```
4. Update `version` and `url` for that dataset only in `inst/config/tidy_data_sources.yml`

## Configuration files

All in `inst/config/`:

- `tidy_data_sources.yml` — version + GitHub Release URL per dataset (user-facing)
- `raw_*_config.yml` — raw source archives, URLs, and periods
- `tidy_*_config.yml` — tidy transformations (filters, derivations, columns)

After changing raw config column structure, run `just update-schemas` to update `inst/schemas/`.

## Dev utilities

Internal functions in `R/dev-utils.R` (load with `devtools::load_all()`):

- `list_archives_periods()` — all archives and available periods
- `list_archive_files("annual_main", "2024-25")` — CSV files inside an archive
- `read_archive_file("annual_main", "2024-25", "main")` — read a specific file
- `compare_schemas("main")` — column changes across periods
- `build_tidy_data("dataset_name")` — build combined parquet for one dataset

## Resources

- [R Packages (2e)](https://r-pkgs.org/)
- [Advanced R](https://adv-r.hadley.nz/)
