---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nhstt

<!-- badges: start -->
[![R-CMD-check](https://github.com/bennettoxford/nhstt/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bennettoxford/nhstt/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/bennettoxford/nhstt/graph/badge.svg)](https://app.codecov.io/gh/bennettoxford/nhstt)
<!-- badges: end -->

> **This package is under active development. Breaking changes are likely.**

`nhstt` provides access to publicly available NHS Talking Therapies reports in a tidy data format.

## Installation

Install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("bennettoxford/nhstt")
```

## Usage

```{r example, eval = FALSE}
library(nhstt)

# Load annual key measures dataset for financial year 2024-25
km_annual <- get_key_measures_annual(periods = "2024-25")

# Load all monthly activity performance datasets
ap_monthly <- get_activity_performance_monthly()
```

## Available NHS Talking Therapies data

### Annual data

```{r datasets-annual, echo = FALSE, message = FALSE}
df_nhstt_reports <- nhstt:::available_nhstt_reports() |>
  dplyr::filter(!stringr::str_detect(title, "metadata")) |>
  dplyr::filter(frequency == "annual") |>
  dplyr::select(
    get_function,
    first_period,
    last_period,
    n_periods,
    version
  ) |>
  dplyr::mutate(get_function = paste0("`", get_function, "()`"))

knitr::kable(
  df_nhstt_reports,
  row.names = FALSE,
  col.names = c(
    "Function",
    "First period",
    "Last period",
    "Count periods",
    "Version"
  ),
  align = c("l", "l", "l", "r", "r")
)
```

### Monthly data

```{r datasets-monthly, echo = FALSE, message = FALSE}
df_nhstt_reports <- nhstt:::available_nhstt_reports() |>
  dplyr::filter(!stringr::str_detect(title, "metadata")) |>
  dplyr::filter(frequency == "monthly") |>
  dplyr::select(
    get_function,
    first_period,
    last_period,
    n_periods,
    version
  ) |>
  dplyr::mutate(get_function = paste0("`", get_function, "()`"))

knitr::kable(
  df_nhstt_reports,
  row.names = FALSE,
  col.names = c(
    "Function",
    "First period",
    "Last period",
    "Count periods",
    "Version"
  ),
  align = c("l", "l", "l", "r", "r")
)
```

### Metadata

```{r datasets-metadata, echo = FALSE, message = FALSE}
df_nhstt_reports <- nhstt:::available_nhstt_reports() |>
  dplyr::filter(stringr::str_detect(title, "metadata")) |>
  dplyr::select(
    get_function,
    first_period,
    last_period,
    n_periods,
    version
  ) |>
  dplyr::mutate(get_function = paste0("`", get_function, "()`")) |>
  dplyr::distinct()

knitr::kable(
  df_nhstt_reports,
  row.names = FALSE,
  col.names = c(
    "Function",
    "First period",
    "Last period",
    "Count periods",
    "Version"
  ),
  align = c("l", "l", "l", "r", "r")
)
```

## For developers

See [DEVELOPERS.md](DEVELOPERS.md).

## Licence

### R package

The `nhstt` package is licensed under the [MIT License](LICENSE.md).

### NHS Talking Therapies data

All NHS Talking Therapies data is Copyright NHS England and licensed under the [Open Government Licence v3.0](https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/).
Contains public sector information licensed under the Open Government Licence v3.0.
