---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# nhstt

<!-- badges: start -->
[![R-CMD-check](https://github.com/bennettoxford/nhstt/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/bennettoxford/nhstt/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/bennettoxford/nhstt/graph/badge.svg)](https://app.codecov.io/gh/bennettoxford/nhstt)
<!-- badges: end -->

`nhstt` provides access to publicly available NHS Talking Therapies reports in a tidy, analysis-ready format.

## Installation

Install the development version from GitHub:

``` r
# install.packages("pak")
pak::pak("bennettoxford/nhstt")
```

## Usage

```{r example, eval = FALSE}
library(nhstt)

# Load annual key measures dataset for financial year 2024-25
km_annual <- get_key_measures_annual(periods = "2024-25")

# Load all monthly activity performance datasets
ap_monthly <- get_activity_performance_monthly()
```

## Available NHS Talking Therapies data

Data can be accessed from R using the `get_*()` functions or downloaded directly using the links in the tables below.
Annual and monthly datasets are also available as [Parquet](https://parquet.apache.org/) files from the GitHub Releases page. 
Metadata datasets link to the raw Excel files published by NHS England.
The _Periods_ column shows the number of reporting periods covered by each dataset.

### Annual data

```{r datasets-annual, echo = FALSE, message = FALSE}
tidy_sources <- yaml::read_yaml(
  system.file("config/tidy_data_sources.yml", package = "nhstt")
)$datasets

tidy_url_lookup <- dplyr::tibble(
  dataset = names(tidy_sources),
  tidy_data = purrr::map_chr(tidy_sources, "url")
) |>
  dplyr::mutate(tidy_data = paste0("[Download](", tidy_data, ")"))

df_nhstt_reports <- nhstt:::available_nhstt_reports() |>
  dplyr::filter(!stringr::str_detect(title, "metadata")) |>
  dplyr::filter(frequency == "annual") |>
  dplyr::left_join(tidy_url_lookup, by = "dataset") |>
  dplyr::select(
    get_function,
    first_period,
    last_period,
    n_periods,
    tidy_data,
    version
  ) |>
  dplyr::mutate(get_function = paste0("`", get_function, "()`"))

knitr::kable(
  df_nhstt_reports,
  row.names = FALSE,
  col.names = c(
    "Function",
    "First period",
    "Last period",
    "Periods",
    "Tidy data",
    "Version"
  ),
  align = c("l", "l", "l", "r", "l", "r")
)
```

### Monthly data

```{r datasets-monthly, echo = FALSE, message = FALSE}
df_nhstt_reports <- nhstt:::available_nhstt_reports() |>
  dplyr::filter(!stringr::str_detect(title, "metadata")) |>
  dplyr::filter(frequency == "monthly") |>
  dplyr::left_join(tidy_url_lookup, by = "dataset") |>
  dplyr::select(
    get_function,
    first_period,
    last_period,
    n_periods,
    tidy_data,
    version
  ) |>
  dplyr::mutate(get_function = paste0("`", get_function, "()`"))

knitr::kable(
  df_nhstt_reports,
  row.names = FALSE,
  col.names = c(
    "Function",
    "First period",
    "Last period",
    "Periods",
    "Tidy data",
    "Version"
  ),
  align = c("l", "l", "l", "r", "l", "r")
)
```

### Metadata

```{r datasets-metadata, echo = FALSE, message = FALSE}
raw_meta_config <- yaml::read_yaml(
  system.file("config/raw_metadata_config.yml", package = "nhstt")
)

raw_url_lookup <- names(raw_meta_config$datasets) |>
  purrr::map(\(ds_name) {
    ds <- raw_meta_config$datasets[[ds_name]]
    if (!is.null(ds$archive)) {
      url <- raw_meta_config$archives[[ds$archive]][[1]]$url
    } else {
      url <- ds$sources[[1]]$url
    }
    dplyr::tibble(get_function = ds$get_function, raw_data = url)
  }) |>
  dplyr::bind_rows() |>
  dplyr::distinct() |>
  dplyr::mutate(raw_data = paste0("[Download](", raw_data, ")"))

df_nhstt_reports <- nhstt:::available_nhstt_reports() |>
  dplyr::filter(stringr::str_detect(title, "metadata")) |>
  dplyr::left_join(raw_url_lookup, by = "get_function") |>
  dplyr::select(
    get_function,
    first_period,
    last_period,
    n_periods,
    raw_data,
    version
  ) |>
  dplyr::mutate(get_function = paste0("`", get_function, "()`")) |>
  dplyr::distinct()

knitr::kable(
  df_nhstt_reports,
  row.names = FALSE,
  col.names = c(
    "Function",
    "First period",
    "Last period",
    "Periods",
    "Raw data",
    "Version"
  ),
  align = c("l", "l", "l", "r", "l", "r")
)
```

## For developers

See [DEVELOPERS.md](DEVELOPERS.md).

## Licence

### R package

The `nhstt` package is licensed under the [MIT License](LICENSE.md).

### NHS Talking Therapies data

All NHS Talking Therapies data is Copyright NHS England and licensed under the [Open Government Licence v3.0](https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/).
Contains public sector information licensed under the Open Government Licence v3.0.
