---
title: "How to use the monthly activity performance data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{how-to-activity-performance-monthly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> **Note:** This vignette is for illustration purposes only and may contain errors. It demonstrates example usage of the dataset and functions, but results should not be interpreted as validated analyses.

This example demonstrates how to explore monthly NHS Talking Therapies reports using the `nhstt` package.
We'll look at changes in monthly measures related to different treatment end codes broken down by individual NHS Talking Therapies providers.

## Setup

We start by loading the necessary packages and downloading the monthly activity performance dataset with `get_activity_performance_monthly()`. This report contains monthly performance indicators for NHS Talking Therapies services across England.

```{r setup, message=FALSE, warning=FALSE}
# Load packages
library(nhstt)
library(ggplot2)
library(dplyr)
library(scales)
library(stringr)

# Get all monthly activity performance reports
activity_performance <- get_activity_performance_monthly()
```


## Select data for analysis

We want to explore changes in the following measures related to referrals that ended:

- **M057**: Count of referrals that ended in the reporting period with an end code of 'Not suitable for IAPT service - no action taken or directed back to referrer' (End code: 10)
- **M058**: Count of referrals that ended in the reporting period with an end code of 'Not suitable for IAPT service - signposted elsewhere with mutual agreement of patient' (End code: 11)
- **M059**: Count of referrals that ended in the reporting period with an end code of 'Discharged by mutual agreement following advice and support' (End code: 12)
- **M060**: Count of referrals that ended in the reporting period with an end code of 'Referred to another therapy service by mutual agreement' (End code: 13)
- **M061**: Count of referrals that ended in the reporting period with an end code of 'Suitable for IAPT service, but patient declined treatment that was offered' (End code: 14)
- **M066**: Count of referrals with an end date in the reporting period - Improving Access to Psychological Therapies care spell end code is 'Mutually agreed completion of treatment’ (End code: 46) 
- **M062**: Count of referrals that ended in the reporting period with an end code of 'Deceased (Seen but not taken on for a course of treatment)' (End code: 17)
- **M063**: Count of referrals that ended in the reporting period with an end code of 'Not known (Seen but not taken on for a course of treatment)' (End code: 95)
- **M066**: Count of referrals with an end date in the reporting period - Improving Access to Psychological Therapies care spell end code is 'Mutually agreed completion of treatment' (End code: 46)
- **M069**: Count of referrals that ended in the reporting period with an end code of 'Deceased (Seen and taken on for a course of treatment)' (End code: 49)
- **M070**: Count of referrals that ended in the reporting period with an end code of 'Not Known (Seen and taken on for a course of treatment)' (End code: 96)
- **M071**: Count of referrals that ended in the reporting period with an invalid end code (Invalid: not in 10, 11, 12, 13, 14, 16, 17, 95, 46, 47, 48, 49, 50, 96)
- **M340**: Count of referrals that ended in the reporting period with an end code of 'Incomplete Assessment (Patient dropped out)' (End code: 16)
- **M341**: Count of referrals that ended in the reporting period with an end code of 'Termination of treatment earlier than patient requested' (End code: 48)
- **M342**: Count of referrals that ended in the reporting period with an end code of 'Not assessed' (End code: 50)
- **M344**: Count of referrals that ended in the reporting period with an end code of 'Termination of treatment earlier than Care Professional planned' (End code: 47)

```{r select-data}
referral_ended_measures <- c(
  "M057", "M058", "M059", "M060", "M061", "M066", "M340", "M062",
  "M063", "M066", "M344", "M341", "M069", "M342", "M070", "M071"
)

selected_activity_performance <- activity_performance |>
  filter(measure_id %in% c(referral_ended_measures)) |>
  filter(group_type == "Provider")
```

## Create figure

Finally, we visualise the trends over time broken down by codes for referrals that ended for each service provider.

```{r fig-example, warning=FALSE, message=FALSE, echo=FALSE, fig.height = 12, fig.width = 10, fig.cap = "Trends for counts of different referrals that ended measures."}

end_dates <- unique(selected_activity_performance$end_date)
fig_date_breaks <- end_dates[seq(4, length(end_dates), by = 4)]

measure_name_to_sentence <- function(x) {
  x |>
    str_remove("^ended_") |>
    str_replace_all("_", " ") |>
    str_to_sentence()
}

fig_selected_activity_performance <- selected_activity_performance |>
  mutate(
measure_name_id = paste0(
  measure_name_to_sentence(measure_name),
      " (",
      measure_id,
      ")"
    )
  ) |>
    select(end_date, org_name2, org_code2, measure_id, measure_name_id, value)
  
levels_order <- fig_selected_activity_performance %>%
    distinct(measure_id, measure_name_id) %>%
    arrange(measure_id) %>%
    pull(measure_name_id)

fig_ref_ended_by_service <- fig_selected_activity_performance |>
  mutate(measure_name_id = factor(measure_name_id, levels = levels_order)) |> 
  ggplot(
    aes(
      y = value,
      x = end_date,
      colour = measure_name_id,
      group = org_code2
    )
  ) +
  geom_line(alpha = 0.2) +
  geom_point(size = .2) +
  labs(
    title = NULL,
    x = "End date of yearly interval",
    y = NULL,
    colour = NULL,
    shape = NULL
  ) +
  scale_x_date(
    # breaks = fig_date_breaks,
    labels = label_date("%b\n%Y")
  ) +
  scale_y_continuous(
    labels = label_number(accuracy = 1),
    limits = c(0, NA)
  ) +
  theme(
    legend.position = "bottom",
    text = element_text(size = 13)
  ) +
  facet_wrap(
    vars(measure_name_id),
    scales = "free_y",
    ncol = 3
  ) +
  guides(colour = guide_legend(nrow = 5, byrow = TRUE))

fig_ref_ended_by_service
```