---
title: "How to use the annual key measures data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{how-to-key-measures-annual}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> **Note:** This vignette is for illustration purposes only and may contain errors. It demonstrates example usage of the dataset and functions, but results should not be interpreted as validated analyses.

This example demonstrates how to explore annual NHS Talking Therapies reports using the `nhstt` package.
We'll look at regional trends in depression referrals and finished treatments to illustrate the basic workflow for accessing and visualising this publicly available data.

## Setup

We start by loading the necessary packages and downloading the annual key measures dataset with `get_key_measures_annual()`. This report contains annual performance indicators for NHS Talking Therapies services across England.

```{r setup, message=FALSE, warning=FALSE}
# Load packages
library(nhstt)
library(ggplot2)
library(dplyr)
library(scales)
library(stringr)

# Get key_measures data
# When running this for the first time it will download and tidy the reports
key_measures <- get_key_measures_annual(
  periods = c("2020-21", "2021-22", "2022-23", "2023-24", "2024-25")
)
```

## Select data for analysis

Here we filter the data to focus on received referrals and finished treatments at the regional level.
This step requires careful exploration of the underlying data structure.

```{r select-data}
# Select data for analysis
selected_measures <- key_measures |>
  filter(org_type == "Commissioning Region") |>
  filter(!org_name == "UNKNOWN") |>
  filter(
    measure_name %in% c("referrals_received", "finished_course_treatment")
  ) |>
  filter(measure_statistic == "count") |>
  filter(variable_a == "Depression") |>
  select(
    start_date,
    end_date,
    org_name,
    variable_a,
    measure_statistic,
    measure_name,
    value
  )

# Tidy data for analysis
selected_measures <- selected_measures |>
  mutate(
    org_name = str_to_title(str_remove(org_name, "\\s*COMMISSIONING REGION$")),
    org_name = str_replace_all(org_name, "\\b(Of|And)\\b", tolower),
    measure_name = factor(
      measure_name,
      levels = c("referrals_received", "finished_course_treatment"),
      labels = c("Referrals received", "Finished course of treatment")
    )
  ) |>
  filter(end_date > "2020-03-31")
```

## Create figure

Finally, we visualise the trends over time for each region.

```{r fig-example, echo=FALSE, fig.height = 12, fig.width = 8, fig.cap = "Trends for counts of referrals received and finished course of treatment for depression broken down by region."}
fig_dep_ref_tx_by_region <- selected_measures |>
  mutate(value = as.numeric(value)) |>
  ggplot(
    aes(
      y = value,
      x = end_date,
      colour = measure_name,
      shape = measure_name
    )
  ) +
  geom_line(alpha = 0.2) +
  geom_point(size = 2) +
  labs(
    title = NULL,
    x = "End date of yearly interval",
    y = NULL,
    colour = NULL,
    shape = NULL
  ) +
  scale_x_date(
    breaks = unique(selected_measures$end_date),
    labels = label_date("%b\n%Y")
  ) +
  scale_y_continuous(
    labels = label_number(accuracy = 1),
    limits = c(0, NA)
  ) +
  theme(
    legend.position = "bottom",
    text = element_text(size = 14)
  ) +
  scale_colour_viridis_d(end = .6) +
  facet_wrap(vars(org_name), ncol = 1)

fig_dep_ref_tx_by_region
```
