---
title: "Exploring monthly changes in treatment end codes"
bibliography: references.bib
link-citations: true
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{how-to-activity-performance-monthly}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> This vignette provides the complete R code used to reproduce the analyses in our blog post: [From public NHS Talking Therapies reports to research insights: Introducing the *nhstt* R package](https://talkingtherapies.opensafely.org/from-public-nhs-talking-therapies-reports-to-research-insights-introducing-the-nhstt-r-package/).

In a [previous blog post](https://talkingtherapies.opensafely.org/introducing-our-series-on-publicly-available-mental-health-data-nhs-talking-therapies-for-anxiety-and-depression/) we introduced the NHS TT data reports that NHS England publishes openly alongside interactive dashboards for exploring key measures directly in a browser.
These datasets cover various measures, including referral activity, waiting times, clinical outcomes, and patient-reported outcome measures.
The value of these public data has been demonstrated by @Clark2018a, who used publicly available NHS TT reports to study between-service variation in clinical outcomes.
They found that organisational factors including waiting times, missed appointments, and the proportion of patients receiving a course of treatment were associated with rates of reliable recovery and improvement.
This illustrates the broader role of public reporting in supporting transparency, benchmarking, and research on variation in routine psychological therapy services [@Clark2018a; @Clark2018b].
The monthly data complement the annual publications by providing more timely signals for monitoring service activity and identifying measures or periods that warrant closer investigation.

The [*nhstt* R package](https://bennettoxford.github.io/nhstt/) is built on top of these publicly available resources and designed to complement them.
It provides an additional point of access to the same data in a tidy, analysis-ready format across all available reporting periods.
In this vignette we use treatment end codes from the monthly activity and performance data as a worked example, but the same approach applies to any of the measures in the dataset.
The examples are descriptive and intended to illustrate what can be done with the data, not to draw conclusions about individual services.

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(nhstt)
library(ggplot2)
library(lubridate)
library(dplyr)
library(tidyr)
library(scales)
library(stringr)
library(paletteer)
library(gt)
```

```{r}
#| label: plot-theme
#| message: false
#| warning: false
# Plot theme: define a custom ggplot2 theme and shared axis scales applied across all figures.
# Custom ggplot theme applied to all figures

theme_nhstt <- function(base_size = 13) {
  theme_classic(base_size = base_size) +
    theme(
      panel.grid.major = element_line(colour = "gray90", linetype = "dotted"),
      strip.text = element_text(face = "bold"),
      strip.background = element_rect(linewidth = 0),
      strip.text.x = element_text(size = base_size, face = "bold"),
      legend.title = element_text(size = base_size),
      legend.text = element_text(size = base_size * 0.95)
    )
}

# Shared axis scales and facet layout used across figures
nhstt_scales <- function(ncol = 2) {
  list(
    scale_x_date(labels = label_date("%b\n%Y")),
    scale_y_continuous(labels = label_number(accuracy = 1), limits = c(0, NA)),
    labs(x = NULL, y = NULL),
    facet_wrap(
      vars(measure_name_id),
      scales = "fixed",
      ncol = ncol,
      axes = "all"
    )
  )
}

theme_set(theme_nhstt())
```

```{r}
#| label: load-data
#| message: false
#| warning: false
# Load data: download monthly activity and performance data and measure metadata.
# Load monthly activity and performance data

df_activity_performance <- get_activity_performance_monthly()
START_DATE_NHSTT <- min(df_activity_performance$start_date)
END_DATE_NHSTT <- max(df_activity_performance$end_date)

# Load measure metadata
df_monthly_metadata <- get_metadata_monthly()
```

```{r}
#| label: define-measures
#| message: false
#| warning: false
# Define measures: select treatment end codes, filter to provider level, and prepare figure labels.
# All five end codes from the seen and taken on for a course of treatment group

referral_ended_measures <- c(
  "M066",
  "M344",
  "M341",
  "M070",
  "M069"
)

# Filter to provider-level data for selected measures
df_activity_performance_selected <- df_activity_performance |>
  filter(measure_id %in% referral_ended_measures) |>
  filter(group_type == "Provider")

# Four most commonly recorded end codes used in the charts (M069 table only)
selected_measures <- c(
  "M066",
  "M070",
  "M341",
  "M344"
)

# Helper to convert measure names to sentence case for figure labels
measure_name_to_sentence <- function(x) {
  x |>
    str_remove("^ended_") |>
    str_replace_all("_", " ") |>
    str_to_sentence()
}

# Two example services highlighted in the deciles chart
selected_service_labels <- c(
  RNUDT = "Service A",
  RXV = "Service B"
)

selected_services <- names(selected_service_labels)

# Facet panel order for figures: follows selected_measures order
top_levels_order <- df_activity_performance_selected |>
  filter(measure_id %in% selected_measures) |>
  distinct(measure_id, measure_name) |>
  mutate(
    measure_name_id = paste0(
      measure_name_to_sentence(measure_name),
      " (",
      measure_id,
      ")"
    )
  ) |>
  arrange(factor(measure_id, levels = selected_measures)) |>
  pull(measure_name_id)
```

## Treatment end codes

When a referral ends in the reporting period, NHS TT services record a reason using a standardised treatment end code.
These codes group referral outcomes into three categories: (1) referred but not seen, (2) seen but not taken on for a course of treatment, and (3) seen and taken on for a course of treatment.
This post focuses on the last category, which covers patients who received two or more sessions, and captures the recorded reason their treatment ended.

The monthly dataset includes counts for each end code, available from `r format(START_DATE_NHSTT, "%B %Y")` to `r format(END_DATE_NHSTT, "%B %Y")`.
The table below shows the five end codes in this group, with total recorded events and the number of services contributing data.
The four most commonly recorded end codes are used in the charts below.
The Deceased end code (M069) is recorded in too few referrals to visualise.
All [descriptions of monthly measures](https://bennettoxford.github.io/nhstt/articles/metadata-monthly-core.html) implemented in the *nhstt* R package can be explored online.

```{r}
#| label: tbl-monthly-end-code-measures
#| message: false
#| warning: false
# Table: summarise total events and number of contributing services for each treatment end code.

df_measure_summary <- df_activity_performance_selected |>
  group_by(measure_id) |>
  summarise(
    total_events = sum(value, na.rm = TRUE),
    n_services = n_distinct(org_code2[!is.na(value)])
  )

df_monthly_metadata |>
  filter(measure_id %in% referral_ended_measures) |>
  select(measure_id, description) |>
  left_join(df_measure_summary, by = "measure_id") |>
  arrange(desc(total_events)) |>
  gt() |>
  tab_style(
    style = list(cell_text(align = "left", v_align = "top", weight = "bold")),
    locations = cells_body(columns = measure_id)
  ) |>
  tab_style(
    style = list(cell_text(align = "left", v_align = "top")),
    locations = cells_body(columns = description)
  ) |>
  fmt_number(columns = total_events, decimals = 0) |>
  cols_label(
    measure_id = "ID",
    description = "Description",
    total_events = "Total events",
    n_services = "NHS TT services"
  ) |>
  tab_caption(
    paste0(
      "End codes for referrals seen and taken on for a course of treatment in the monthly NHS TT dataset (",
      format(START_DATE_NHSTT, "%B %Y"),
      " to ",
      format(END_DATE_NHSTT, "%B %Y"),
      "), with total recorded events and number of services contributing data."
    )
  )
```

## Monthly trends and variation in treatment end codes

Across `r length(unique(df_activity_performance$reporting_period))` reporting periods, these monthly counts make it possible to look beyond annual totals and examine whether patterns in treatment end codes are stable or change over time.
We first show service-level trends across all services, then use decile bands to place two example services in context.

### Trends across all services

The figure below shows the raw monthly counts for the four most commonly recorded end codes across all NHS TT services.
Each line represents one service, showing both the scale of between-service variation and whether it is stable over time.
Because these are counts rather than rates, differences between services will partly reflect differences in service size.

```{r}
#| label: fig-monthly-end-code-measures
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 9
#| fig-cap: "End codes for referrals seen and taken on for a course of treatment in the monthly NHS TT dataset: monthly counts across all NHS TT services for the four most commonly recorded end codes."
# Figure: monthly counts for the four most commonly recorded end codes across all NHS TT services.

fig_selected_activity_performance <- df_activity_performance_selected |>
  filter(measure_id %in% selected_measures) |>
  mutate(
    measure_name_id = factor(
      paste0(measure_name_to_sentence(measure_name), " (", measure_id, ")"),
      levels = top_levels_order
    )
  ) |>
  select(end_date, org_name2, org_code2, measure_id, measure_name_id, value)

fig_selected_activity_performance |>
  ggplot(
    aes(
      y = value,
      x = end_date,
      colour = measure_name_id,
      group = org_code2
    )
  ) +
  geom_line(alpha = 0.2) +
  geom_point(size = 0.2) +
  scale_colour_brewer(palette = "Dark2") +
  nhstt_scales() +
  theme(legend.position = "none") +
  guides(colour = guide_legend(nrow = 5, byrow = TRUE))
```

Across all four end codes, most services record counts within a similar range and follow broadly consistent trends over the reporting period.
A small number of services stand out as consistent outliers, particularly for referrals that ended with *Completed* (M066), *Before care professional planned* (M344), and *Unknown treated* (M070).

### Comparing services with decile bands

The figure below uses decile bands to place the two example services in the context of the full distribution across all NHS TT services.
Read more on why we think [deciles charts are a useful tool to communicate variation](https://www.bennett.ox.ac.uk/blog/2019/04/communicating-variation-in-prescribing-why-we-use-deciles/) in a previous blog we wrote.
The shaded bands show the distribution from the 10th to the 90th percentile in decile steps, with darker shading closer to the median.
A service outside the bands falls in the top or bottom 10% for that month.
The width of the bands reflects how much variation there is between services, with wider bands indicating greater spread.
These charts summarise activity patterns and are not intended as rankings of service quality.

```{r}
#| label: fig-monthly-end-code-measures-deciles
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 10
#| fig-cap: "End codes for referrals seen and taken on for a course of treatment in the monthly NHS TT dataset: decile charts across all NHS TT services for the four most commonly recorded end codes. Shaded bands show the 10th to 90th percentile range in decile steps. The dark line shows the median. Two example services are shown as coloured lines."
# Figure: decile chart showing the distribution across all services, with two example services highlighted.
# Calculate percentiles across all services for each reporting period

df_deciles <- df_activity_performance_selected |>
  filter(measure_id %in% selected_measures) |>
  summarise_percentiles(
    period_col = end_date,
    value_col = value,
    group_by = measure_id
  ) |>
  left_join(
    df_activity_performance_selected |>
      distinct(measure_id, measure_name) |>
      mutate(
        measure_name_id = factor(
          paste0(measure_name_to_sentence(measure_name), " (", measure_id, ")"),
          levels = top_levels_order
        )
      ),
    by = "measure_id"
  ) |>
  pivot_wider(names_from = percentile, values_from = value, names_prefix = "p")

# Prepare data for the two highlighted services
df_top <- df_activity_performance_selected |>
  filter(measure_id %in% selected_measures) |>
  filter(org_code2 %in% selected_services) |>
  select(-org_name2) |>
  mutate(
    service_label = recode(org_code2, !!!selected_service_labels),
    service_label = factor(
      service_label,
      levels = unname(selected_service_labels)
    ),
    measure_name_id = factor(
      paste0(measure_name_to_sentence(measure_name), " (", measure_id, ")"),
      levels = top_levels_order
    )
  )

# Visual style for decile ribbons, median line, and service lines
ribbon_palette <- setNames(
  gray(seq(0.7, 0.5, length.out = 4)),
  c("10th-90th", "20th-80th", "30th-70th", "40th-60th")
)

ribbon_alpha <- 0.2

median_colour <- paletteer_d("RColorBrewer::RdGy")[10]

service_colours <- setNames(
  c(
    paletteer_d("RColorBrewer::Dark2")[1],
    paletteer_d("RColorBrewer::Dark2")[4]
  ),
  unname(selected_service_labels)
)

# Build plot
ggplot() +
  geom_ribbon(
    data = df_deciles,
    aes(x = end_date, ymin = p10, ymax = p90, fill = "10th-90th"),
    alpha = ribbon_alpha
  ) +
  geom_ribbon(
    data = df_deciles,
    aes(x = end_date, ymin = p20, ymax = p80, fill = "20th-80th"),
    alpha = ribbon_alpha
  ) +
  geom_ribbon(
    data = df_deciles,
    aes(x = end_date, ymin = p30, ymax = p70, fill = "30th-70th"),
    alpha = ribbon_alpha
  ) +
  geom_ribbon(
    data = df_deciles,
    aes(x = end_date, ymin = p40, ymax = p60, fill = "40th-60th"),
    alpha = ribbon_alpha
  ) +
  geom_line(
    data = df_deciles,
    aes(x = end_date, y = p50),
    colour = median_colour,
    linewidth = 0.6,
    show.legend = FALSE
  ) +
  geom_line(
    data = df_top,
    aes(x = end_date, y = value, colour = service_label, group = service_label),
    alpha = 0.5
  ) +
  geom_point(
    data = df_top,
    aes(
      x = end_date,
      y = value,
      shape = service_label,
      colour = service_label,
      group = service_label
    ),
  ) +
  scale_colour_manual(name = "NHS TT service:", values = service_colours) +
  scale_shape_manual(
    name = "NHS TT service:",
    values = c("Service A" = 16, "Service B" = 17)
  ) +
  scale_fill_manual(
    values = ribbon_palette,
    breaks = c("10th-90th", "20th-80th", "30th-70th", "40th-60th")
  ) +
  nhstt_scales() +
  labs(
    colour = "NHS TT service:",
    shape = "NHS TT service:",
    fill = "Decile range:"
  ) +
  guides(
    colour = guide_legend(order = 1, nrow = 1),
    shape = guide_legend(order = 1, nrow = 1),
    fill = guide_legend(order = 2, nrow = 1)
  ) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical"
  ) +
  facet_wrap(
    vars(measure_name_id),
    scales = "free_y",
    ncol = 2,
    axes = "all"
  )
```

*Service A* sits above the median for completed referrals (M066) and tracks closer to the median for the remaining end codes.
*Service B* records consistently high counts of completed referrals, sitting in the top 10% of services throughout the reporting period, and also starts above the median for end codes before the patient requested (M341) and care professional planned (M344), trending closer towards the median over the reporting period.

## Try it yourself

The *nhstt* package makes it easier to access publicly available NHS TT data, is free to use, and is updated regularly as NHS England publishes new reports.
It provides access to each dataset in a tidy, analysis-ready format, with no need to locate or parse the underlying files.
Whether you are interested in trends in referral activity, variation in waiting times across services, or changes in clinical outcomes over time, the data are there to be explored.
We recommend checking the NHS England [data quality notes](https://digital.nhs.uk/binaries/content/assets/website-assets/data-and-information/datasets/nhs-talking-therapies/nhs_talking_therapies_dq_note-260327.xlsx) before using these data, as they describe known issues affecting specific reporting periods or measures.

If you use the package for a new analysis or have ideas for how it could be extended, we would love to hear about it.
For questions or to report an issue, please [open an issue on GitHub](https://github.com/bennettoxford/nhstt/issues) or get in touch with [Milan Wiedemann](https://www.phc.ox.ac.uk/team/milan-wiedemann).

## References
