---
title: "Quarterly measures and breakdowns"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quarterly measures and breakdowns}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

> **Note:** The quarterly metadata comes from the most recent monthly metadata file. Older quarterly periods might have fewer, or different, measures.

```{r setup, include = FALSE}
library(nhstt)
library(dplyr)
library(stringr)
library(reactable)

metadata_monthly <- get_metadata_monthly()

# Metadata for quarterly measures
metadata_quarterly_measures <- metadata_monthly |>
  filter(
    measure_type != "Variable",
    frequency %in% c("Quarterly", "Monthly/Quarterly")
  ) |>
  select(measure_id, description)

# Metadata for quarterly breakdowns, parsed from the measure name
metadata_quarterly_variables <- metadata_monthly |>
  filter(measure_type == "Variable") |>
  mutate(
    # str_squish because some levels have a double space after the label
    variable_type = str_squish(
      str_extract(measure_name, "(?<=Variable Type: )[^\r\n]+")
    ),
    variable_a = str_squish(
      str_extract(measure_name, "(?<=Variable A: )[^\r\n]+")
    ),
    variable_b = str_squish(
      str_extract(measure_name, "(?<=Variable B: )[^\r\n]+")
    )
  ) |>
  select(variable_type, variable_a, variable_b, description)

# Add a border when the variable type changes
border_style <- JS(
  "function(rowInfo, column, state) {
  if (rowInfo.viewIndex > 0) {
    const prevRow = state.pageRows[rowInfo.viewIndex - 1]
    if (prevRow && rowInfo.values['variable_type'] !== prevRow['variable_type']) {
      return { borderTop: '1px solid #333' }
    }
  }
}"
)
```

## Measures

In total there are **`r length(unique(metadata_quarterly_measures$measure_id))`** measures that the metadata lists as quarterly:

```{r tab-measures-quarterly, echo=FALSE}
reactable(
  metadata_quarterly_measures,
  searchable = TRUE,
  sortable = FALSE,
  defaultPageSize = 10,
  columns = list(
    measure_id = colDef(
      name = "Measure ID and description",
      cell = function(value, index) {
        description <- metadata_quarterly_measures$description[index]
        htmltools::tags$div(
          htmltools::tags$code(
            style = "color: #2A788EFF;",
            value
          ),
          ": ",
          description
        )
      }
    ),
    description = colDef(show = FALSE, searchable = TRUE)
  ),
  outlined = TRUE,
  striped = TRUE
)
```

The quarterly data also include the waiting time measures `M347` to `M350` and `M352` to `M355`, which the metadata lists as monthly measures.
See `vignette("metadata-monthly-core")` for their descriptions.

## Breakdowns

### Organisational

The `org_type` column shows the organisation level of each row (e.g., England, CommissioningRegion, ICB, SubICB, Provider, SubICB-Provider).
The `icb_code`, `sub_icb_code` and `provider_code` columns (and their name columns) are filled for the levels that apply to the row and `NA` otherwise.

### Demographic and clinical

These three columns provide further options for breakdowns:

- `variable_type`: The type of variable by which the data are categorised.
- `variable_a`: The high-level variable values for the categories.
- `variable_b`: The sub-category, where applicable. `NA` if not applicable.

The data also include a `Total` variable type with all records.
The following table shows the other **`r nrow(metadata_quarterly_variables)`** breakdowns:

```{r tab-variables-quarterly, echo=FALSE}
# Following the merge cells example at:
# https://glin.github.io/reactable/articles/cookbook/cookbook.html#merge-cells
reactable(
  metadata_quarterly_variables,
  searchable = TRUE,
  sortable = FALSE,
  defaultPageSize = 10,
  columns = list(
    variable_type = colDef(
      name = "Variable type",
      style = JS(
        "function(rowInfo, column, state) {
        const firstSorted = state.sorted[0]
        // Merge cells if unsorted or sorting by variable_type
        if (!firstSorted || firstSorted.id === 'variable_type') {
          const prevRow = state.pageRows[rowInfo.viewIndex - 1]
          if (prevRow && rowInfo.values['variable_type'] === prevRow['variable_type']) {
            return { visibility: 'hidden' }
          }
        }
        // Add border when variable_type changes
        if (rowInfo.viewIndex > 0) {
          const prevRow = state.pageRows[rowInfo.viewIndex - 1]
          if (prevRow && rowInfo.values['variable_type'] !== prevRow['variable_type']) {
            return { borderTop: '1px solid #333' }
          }
        }
      }"
      )
    ),
    variable_a = colDef(name = "Variable A", style = border_style),
    variable_b = colDef(name = "Variable B", style = border_style),
    description = colDef(name = "Description", style = border_style)
  ),
  outlined = TRUE,
  striped = TRUE
)
```
